\documentclass{article}
\usepackage{parskip}
\title{CSC 429 Assignment 1}
\author{Alexander Williams}
\date{\today}
\begin{document}
\maketitle
\section*{Question 1}
\subsection*{a)}
Given a block length of $\ell=1$, and a key of $k_0,k_1,...,k_{\ell-1}$,
the encryption of the $i^{th}$ character in a block where $0\leq{}i\leq\ell-1$ is:
$$
c_i=k_i+m_i+j
$$
The corresponding decryption function will be written as:
$$
m_i=c_i-k_i-j
$$

\subsection*{b)}
First, find the key length with the index of coincidence.

Since each character in a block is shifted by a different amount, we only want
to calculate the IC of all the $i^{th}$ characters in a block together. For
example, We take all the $1^{st}$ characters in a block, and find their IC.
Then the $2^{nd}$, then $3^{rd}$, up to the $\ell^{th}$ character, and we take 
the average to find the overall IC for a key of that length.

We do this for all possible key length, and we take the length with the closest
IC to $0.065$, we try to minimize the function
$|\sum_{0}^{25}{q_i^2}=0.065|$,
where $q_i$ is the the probability that a given character in the text is the
$i^{th}$ character in the alphabet

Since the $j^{th}$ block increments the value of each character by $j$, we'll have
to subtract $j$ from the value of each character before taking the IC.

Once we have the block length, we have to find the key.

Once again, we'll take the $i^{th}$ character of each block, subtract $j$, and
this time we want to find the key character $k$ that minimizes
$|\sum_{0}^{25}p_iq_{(i-k)mod26}-0.065|$, where $q_{(i-k)mod26}$ is the probability 
of any character in the text is the $i-k^{th}$ character in the alphabet,
and $p_i$ is the chance any character in the english language is the $i^{th}$
character in the alphabet.

Once we've done this to get all $\ell$ characters of the key, we can use the 
earlier mentioned decryption function to get back the original message.

\subsection*{c)}
Running our ciphertext through a python program I made for this assignment, I got
a key length of 6, and the key was $(2,8,15,7,4,14)$ (C I P H E R).

Once decoded, the message reads:

AGEDTWENTYSIXVIGENEREWASSENTTOROMEONADIPLOMATICMISSION
ITWASHERETHATHEBECAMEACQUAINTEDWITHTHEWRITINGSOFALBERTITRITHEMIUSANDPORTAANDHISINTERESTIN
CRYPTOGRAPHYWASIGNITEDFORMANYYEARSCRYPTOGRAPHYWASNOTHINGMORETHANATOOL
THATHELPEDHIMHISDIPLOMATICWORKBUTATTHEAGEOFTHIRTYNINEVIGENEREDECIDEDTHATHE
HADAMASSEDENOUGHMONEYTOBEABLETOABANDONHISCAREERANDCONCENTRATEONALIFEOF
STUDYITWASONLYTHENTHATHEBEGANRESEARCHINTOANEWCIPHER

\end{document}