\documentclass{article}
\usepackage{parskip}
\title{CSC 429 Assignment 1}
\author{Alexander Williams}
\date{\today}
\begin{document}
\maketitle
\section*{Question 6}

\subsection*{a)}

Proof that $negl_3(n)=negl_1(n)+negl_2(n)$ is negligible

For a function to be negligible, there must exist a value $N$
such that for all $n\geq{}N$, $negl(n)\leq\frac{1}{q(n)}$, where 
$q(n)$ is any polynomial function.

By definition, a polynomial function multiplied by a constant
is still a polynomial function. This means $2q(n)$ is polynomial,
given the fact that $q(n)$ is polynomial. Let $q'(n)=2q(n)$

It follows that there exists both an $N_1$ such that for all $n\geq{}N_1$, 
$negl_1(n)\leq{}\frac{1}{q'(n)}$, and an $N_2$ such that for all $n\geq{}N_2$,
$negl_2(n)\leq{}\frac{1}{q'(n)}$.

Let $max(N_1, N_2)=N'$.

Since for all $n\geq{}N'$, $negl_2(n)\leq{}\frac{m}{q'(n)}$ and $negl_1(n)\leq{}\frac{1}{q'(n)}$,
it follows that $negl_1+negl_2 < \frac{2}{q'(n)}$.

Since $q'(n)=2q(n)$, we can simplify this expression.

$$
\frac{2}{q'(n)}=\frac{2}{2q(n)}=\frac{1}{q(n)}
$$

This shows that $negl_1(n)+negl_2(n)$ is negligible.

Therefore, $negl_3(n)=negl_1(n)+negl_2(n)$ is negligible.



\subsection*{b)}

Proof that $negl_4=p(n)*negl_1(n)$, where $p(n)$ is a polynomial function,
is negligible.

For a function to be negligible, there must exist a value $N$
such that for all $n\geq{}N$, $negl(n)\leq\frac{1}{q(n)}$, where 
$q(n)$ is any polynomial function.

By definition, a polynomial function multiplied by a polynomial function
is a polynomial function. This means that $p(n)*p(n)=p^2(n)$ is polynomial.

Let $q(n)=p^2(n)$.

This means that for all $n\geq{}N$:

$$
p(n)*negl_1(n)=p(n)*\frac{1}{q(n)}
$$
$$
=\frac{p(n)}{p^2(n)}
$$
$$
=\frac{1}{p(n)}
$$

Since $p(n)$ is a polynomial function, $p(n)*negl_1(n)$ is negligible.



\subsection*{c)}

show that $f(n)=\sum_{j=1}^{n}negl_j(n)$ is negligible

As shown earlier in part a, the sum of two negligible functions is still negligible.

$\sum_{j=1}^{n}negl_j(n)$ can be re-written as:

$$
negl_1(n)+negl_2(n)+...+negl_n(n)
$$
Since the sum of two negligible functions is negligible, we can simplify
by combining $negl_1,negl_2$.
$$
=negl_{s1}(n)+negl_3(n)+...+negl_n(n)
$$
We can then sum together $negl_{s1},negl_3$
$$
=negl_{s2}(n)+negl_4(n)+...+negl_n(n)
$$
$$
...
$$
$$
=negl_{s(n-2)}(n)+negl_n(n)
$$
$$
=negl_{s(n-1)}(n)
$$

This shows that:
$$
\sum_{j=1}^{n}negl_j(n)=negl_{s(n-1)}(n)
$$

Thus, $\sum_{j=1}^{n}negl_j(n)$ is negligible.

\end{document}