---
title: "Set_10"
author: "Alexander Williams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Contingency Tables

So far we've been looking at data from continuous measurements. Contingency tables are for comparing discrete values, things like yes/no answers, or where there's only a handful of possible values.

___

## Example: Demographic Study in BC

a random sample of 1436 women over 30 who were married at least once
```{r}
demog<-matrix(c(550,681,61,144),nrow=2,ncol=2,
              dimnames=list(c("College","No College"),
                            c("Married Once","    Married > 1")
                            ))

demog
colSums(demog)
rowSums((demog))
```

Is there any association between marital status and education?

This is a $2\times2$ contingency table, two factors with two levels. We can use a chi-squared test to answer this question

$$\chi^2=\sum_{i=1}^r\sum_{j=1}^c\frac{(O_{ij}-E_{ij)^2}}{E_{ij}}$$
**Important:** all values in the table must be greater than 5, otherwise another test needs to be used.

Where $O_{ij}$ is what you observe, and $E_{ij}$ is the expected number assuming independence. In our case, $O_{12}$ is first row, second column, married with no college. Our expected value is from the following equation:
$$E_{ij}=\frac{N_{i*}N_{*j}}{N}$$
Lets look at the table of the expected values
```{r}
chisq.test(demog,correct=FALSE)$expected
```

We sum everything together in the equation shown earlier, and we get $\chi^2=16$, in the distribution 
$$\chi^2_{(r-1)(c-1)}=\chi^2_{1}$$
```{r}
chisq.test(demog, correct=FALSE)
```
What is that `correct=FALSE`? this test is only an approximation, it gets more accurate as we get more datapoints. We can apply this correction to account for this.
```{r}
chisq.test(demog)
```

With such small p-values, we have evidence to reject the null hypothesis of no correlation between marriages and education. Let's look at the actual and expected values again
```{r}
demog
chisq.test(demog,correct=FALSE)$expected
```

We see less with a College education married only once, and less people married multiple times with no college education

___

## Example: Diet and Prostate Cancer

6272 Swedish men followed for 30 years to see if there's an association between fish in their diet and prostate cancer.
```{r}
fish <- read.table(file='~/Documents/uni/STAT359/data/fish_diet.txt',
                   header=TRUE,sep="")
attach(fish)
names(fish)
table(fish)
```
We have the data, but we need to save it as the table in order to do the test.

For this example, we won't use the correction even though that gets us a lightly less accurate result.

```{r}
fish.table <- as.matrix(table(fish))
chisq.test(fish.table, correct=FALSE)
```

the p-value is 0.2985, there's no evidence that the amount of fish in your diet affects your chance of getting cancer.

Just for fun, let's compare the observed table and the expected table.
```{r}
fish.table
chisq.test(fish.table,correct=FALSE)$expected
```
You can visible see that for the most part, expected and observed values are very close to each other.









