---
title: "Set 4"
author: "Alexander Williams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Skewness and Kurtosis

variability associated with random variable X is defined by:
$$VAR[X]=E[(X-E[X])^2]$$

The skewness of variable X is defined by:
$$Skew[X]=E[(\frac{X-\mu}{\sigma})^3]$$
if Skew[X] >0, than the right tail is longer.

```{r}
# load skew-normal package
library(sn)

# positively skewed example
x<-seq(-4,4,0.1)
y<-dsn(x,alpha=5)
plot(x,y,type='l',col='green',lwd=3)
title('Density of a right-skewed distribution')
```
If Skew[X] < 0, the left tail is longer
```{r}
x<-seq(-4,4,0.1)
y<-dsn(x,alpha=-5)
plot(x,y,type='l',col='green',lwd=3)
title('Density of a left-skewed distribution')
```
If Skew[X] = 0, the distribution is symmetric.

The Skewness can be estimated with the following equation
$$\hat{Skew}=\frac{\sum_{i=1}^n(y_i-\bar{y})^3}{ns^3}$$
we can also use this to create confidence intervals on skewness

___

# Example: Skewness
```{r}
stereograms<-read.table(file="~/Documents/uni/STAT359/data/stereograms.txt",
                         sep="",
                         header=TRUE)
time.NV<-stereograms$fusion_time[stereograms$group=='NV']
time.VV<-stereograms$fusion_time[stereograms$group=='VV']
boxplot(time.NV,time.VV,col='green',names=c('Time - NV','Time - VV'))
title('Stereogram Fusion Times')

qqnorm(time.NV,main='QQ-Plot: No/Verbal Information')
qqnorm(time.VV,main='QQ-Plot: Verbal/Visual Information')
```



```{r}
# function to compute skewness
skew <-function(x)
{
  m3 <- sum((x-mean(x))^3)/length(x)
  s3<-sqrt(var(x))^3
  m3/s3
}
```

```{r}
skew.hat.NV <- skew(time.NV)
skew.hat.NV
skew.hat.VV<-skew(time.VV)
skew.hat.VV
```
 Now lets useconstruct the 95% confidence interval for skewness

```{r}
x<-time.NV ## data for bootstrapping
B<-15000
# Doing all samples at once via vectorization
#time.NV
x.boot<-matrix(data=sample(x=x,size=B*length(x),replace=TRUE),
               nrow=length(x),
               ncol=B)
skew.boot.sampled<-apply(x.boot,# applies function to whole matrix
                         2,     # 2 means to each column
                         skew)  # function to be used
boot.interval<-quantile(skew.boot.sampled,probs=c(0.025,0.975))
skew.hat.NV
boot.interval
# the above shows that 95% of the time, the skew will be between 0.57 and 3.44, positively skewed

#Time.VV
x.boot<-matrix(data=sample(x=x,size=B*length(x),replace=TRUE),nrow=length(x),ncol=B)
skew.boot.sampled<-apply(x.boot,2,skew)
boot.interval<-quantile(skew.boot.sampled,probs=c(0.025,0.975))
skew.hat.VV
boot.interval
```
Neither confidence interval contains 0, so it is likely both distributions are skewed to the right. 

We also note that the skewness confidence interval overlaps quite a bit, the interval of the VV group is entirely contained in the NV group confidence interval. This means there is no evidence that the skewness of one group is any different from the other.

___

# Example - Slalom times

```{r}
slalom2014<-read.table(file="~/Documents/uni/STAT359/data/slalom2014.txt",
                         sep="",
                         header=TRUE)
names(slalom2014)
attach(slalom2014)

summary(Time_sec)
sqrt(var(Time_sec))

Time.skew.est<-skew(Time_sec)
Time.skew.est

boxplot(Time_sec, col='green')
title('Giant Slalom Times')

qqnorm(Time_sec,main='QQ Plot Normal - Normal')
```
This data is very clearly skewed, as is common in event time data.

Now, lets get the confidence interval on the times.
```{r}
x<-Time_sec ## data for bootstrapping
B<-15000
x.boot<-matrix(data=sample(x=x,size=B*length(x),
                           replace=TRUE),
               nrow=length(x),
               ncol=B)

skew.boot.sampled<-apply(x.boot, # Data to sample from
                         2,      # 2 means apply to rows, 1 means to columns
                         skew)   # function to apply
boot.interval<-quantile(skew.boot.sampled,probs=c(0.025,0.975))

hist(skew.boot.sampled, 
     main='Empirical Distribution for Skew.hat',
     xlab='Sampled Values')

# Create red line showing observed median
abline(v=Time.skew.est,
       col='red') ## arguments can be a and b, h, or v

Time.skew.est
```


# Kurtosis

Measure of fourth moment of distribution, measures weight of tails and how pointed/flat the peak is relative to a normal distribution.
$$Kurt[X]=E[(\frac{X-\mu}{\sigma})^4]-3$$
We can calculate sample kurtosis with:
$$\hat{Kurt}=\frac{\sum_{i=1}^n(y_i-\bar{y})^4}{ns^4}-3$$

Positive kurtosis means heavier tails, like a t-distribution
```{r}
x<-seq(-5,5,.01)
y.norm<-dnorm(x,mean=0,sd=1)
y.t1<-dt(x,df=1)
y.t2<-dt(x,df=2)
y.t3<-dt(x,df=3)
y.t4<-dt(x,df=4)
y.t10<-dt(x,df=10)

# set up the plot area
plot(c(min(x),max(x)),c(min(y.norm,y.t1,y.t2,y.t3,y.t4,y.t10),max(y.norm,y.t1,y.t2,y.t3,y.t4,y.t10)),type="n",
xlab="x",ylab="Density function: f(x)")
title("Comparing the Normal and t-distribution")
lines(x,y.norm)
lines(x,y.t10,lty=2,col="red")
lines(x,y.t4,lty=3,col="red")
lines(x,y.t3,lty=4,col="red")
lines(x,y.t2,lty=5,col="red")
lines(x,y.t1,lty=6,col="red")
```

Negative kurtosis means lighter tails, and often a flatter shape. A uniform distribution is a good example

Function to estimate kurtosis
```{r}
kurtosis<-function(x)
  {
  m4<-sum((x-mean(x))^4)/length(x)
  s4<-var(x)^2
  m4/s4 - 3
  }
```

What values to you get with a normal distribution? We expect 0
```{r}
x.norm <- rnorm(1000)
x.norm.kurt <- kurtosis(x.norm)
x.norm.kurt

# We can bootstrap it to get a confidence interval for the kurtosis
x <- x.norm # data to bootstrap
B <- 10000 # times to bootstrap

x.boot <- matrix(data=sample(x=x,size=B*length(x),replace=TRUE),
                 nrow=length(x),
                 ncol=B)
x.boot.kurt <- apply(x.boot,
                     2,
                     kurtosis)
boot.interval <- quantile(x.boot.kurt,
                          probs=c(0.025, 0.975))
hist(x.boot.kurt, main='Empirical Distribution for kurtosis, normal distribution',xlab='Sampled Values')
abline(v=x.norm.kurt, col='red') ## arguments can be a and b, h, or v
```

The value varies every time it's run, but should be very close to zero on average. Lets do the same steps again, but with a t-distribution
```{r}
x.t <- rt(1000, df=4)
x.t.kurt <- kurtosis(x.t)
x.t.kurt

# bootstrap
x <- x.t
B <- 10000 # times to bootstrap

x.boot <- matrix(data=sample(x=x,size=B*length(x),replace=TRUE),
                 nrow=length(x),
                 ncol=B)
x.boot.kurt <- apply(x.boot,
                     2,
                     kurtosis)
boot.interval <- quantile(x.boot.kurt,
                          probs=c(0.025, 0.975))
hist(x.boot.kurt, main='Empirical Distribution for kurtosis, t-distribution',xlab='Sampled Values')
abline(v=x.t.kurt, col='red') ## arguments can be a and b, h, or v
```

This is much higher than the normal distribution, implying excess kurtosis compared to the normal distribution.

What a higher kurtosis means is that data is less likely to be near the mean. Greater kurtosis means it's more likely to see an extreme result

___

# Example: stereograms
```{r}
kurt.time.NV <- kurtosis(time.NV)
kurt.time.VV <- kurtosis(time.VV)
kurt.time.NV
kurt.time.VV
# Is that a very large kurtosis, or is there too much uncertainly due to small sample size? lets bootstrap
x <- time.NV
B <- 10000 # times to bootstrap

x.boot <- matrix(data=sample(x=x,size=B*length(x),replace=TRUE),
                 nrow=length(x),
                 ncol=B)
x.boot.kurt <- apply(x.boot,
                     2,
                     kurtosis)
boot.interval <- quantile(x.boot.kurt,
                          probs=c(0.025, 0.975))
hist(x.boot.kurt, main='Empirical Distribution for kurtosis, time.NV',xlab='Sampled Values')
abline(v=x.t.kurt, col='red') ## arguments can be a and b, h, or v

boot.interval
```

It's hard to say, there's a lot of uncertainty. The expected values appears to be around 10. but the confidence interval does include 0, so we can't reject the possibility that the true kurtosis is normal.

```{r}
x <- time.VV
B <- 10000 # times to bootstrap

x.boot <- matrix(data=sample(x=x,size=B*length(x),replace=TRUE),
                 nrow=length(x),
                 ncol=B)
x.boot.kurt <- apply(x.boot,
                     2,
                     kurtosis)
boot.interval <- quantile(x.boot.kurt,
                          probs=c(0.025, 0.975))
hist(x.boot.kurt, main='Empirical Distribution for kurtosis, time.VV',xlab='Sampled Values')
abline(v=x.t.kurt, col='red') ## arguments can be a and b, h, or v

boot.interval
```

The interval is much smaller, and also contains 0 so we have no evidence that the kurtosis is not normal
