---
title: "Set_7"
author: "Alexander Williams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```
# Mann-Whitney Test, a Nonparametric Two Sample Procedure

In a two-sample setting, where the data isn't paired, the Mann-Whitney test is a non-parametric alternative to the t-test without any assumptions. This test again uses the idea of replacing the actual data values with their ranks, like the signed-rank test

we group both samples together, ranking all $m+n$ samples. Our test statistic $T_Y$ is then the sum of all ranks from the first sample

Under the null hypothesis that both samples are from the same distribution, it's equally likely that either sample will have larger ranks assigned to it.

each of the $_{m+n}C_m$ assignments of ranks to the $y$ sample is equally likely, with probability $(_{m+n}C_m)^{-1}$


___

# Example: Iron retention

Determining if Fe2+ or Fe3+ are retained differently to see what makes a better dietary supplement

```{r}
iron <- read.table(file = '~/Documents/uni/STAT359/data/iron.txt',
                   sep = "",
                   header=TRUE)
kable(iron)
qqnorm(iron$Fe2)
qqnorm(iron$Fe3)

```

both are bowed, appear non-normal, positively skewed. For the t-test, we applied a log transform first. Being normal doesn't matter for the Mann-Whitney test since we make no assumptions, so we do the test directly.

```{r}
wilcox.test(iron$Fe3, iron$Fe2, paired=FALSE)
```

with a p-value of 0.37, there is not enough evidence to say the different forms of iron are retained differently.











