---
title: "Assignment 6"
author: "Alexander Williams"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# 1

```{r}
geriatric <- read.table(file='~/Documents/uni/STAT359/data/geriatric.txt',
                        header=TRUE, sep='')
```

## a)

```{r}
mod1 <-glm(Fall~exp(Int)+exp(Sex)+exp(BI)+exp(SI),
           data=geriatric, family=poisson)
summary(mod1)
```

\newpage

## b)

```{r}
deviance(mod1)
mod1$df.residual
1-pchisq(deviance(mod1), mod1$df.residual)
```

The p-value is $\approx0.035$, meaning there is evidence that this model is adequate for this data.

## c)

```{r}
plot(mod1$residuals)
```

While most points stay within the range $[-1, 2]$, point 34 does appear greater than the rest.

```{r}
mod1$residuals[34]
```

While this point does stand out, it isn't far out of the range of the other values, so it may not have a significant impact.

## d)

```{r}
mod2 <- update(mod1, .~. - exp(Sex))
anova(mod1, mod2, test="Chi")
```

In this case, we have no evidence that a model without sex peforms any better than the model that includes it. 

## e)

```{r}
summary(mod2)
confint(mod2, "exp(Int)")
```

According to this confidence interval, intervention will reduce the average number of falls by between 0.4 and 0.7.

## f) 

 
We have reason to believe that aerobic education and exercise do lead to a reduction of falls, even after taking into account both balance and strength. 






















